Contracts for Haskell

### 2011-06-22 contracts.tex
pdf update

\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{fullpage}

@@

Semantics of contract satisfaction (exactly the same as in the POPL paper):
\begin{center}
\begin{array}{rcl}
  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,\unr\})\\
  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
  e \in \any &\iff& True  
\end{array}
\end{center}

@@

>>>>

\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{graphicx}

@@

Semantics of contract satisfaction (exactly the same as in the POPL paper):
\begin{center}
\begin{array}{rcl}
  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
  e \in \any &\iff& True  
\end{array}
\end{center}

@@

A stupid fix was to provide defining equations for $multp$ (the same
as for $mult$). Indeed, we could prove that $mult$ satisfied its
contract, but it was trivially true...

Current idea is to unfold the definition a fixed number of times.

\section{De-appification and removed CFness}
\subsection{De-appification}
Instead of always using the $app$ function when doing function
application, we now use directly the function name if it is fully
applied. For example, if $f$ is of arity 2, $f x y$ is now translated
to $f(x,y)$ instead of $app(app(f,x),y)$.  To deal with partially
applied functions, we can use $f\_ptr$ and some extra axioms.

This transformation makes things much faster for equinox, especially
given that in most cases, function are totally applied.

\subsection{Removing CFness}
In the POPL paper, CF was introduced in the contract satisfaction
semantics because it was required in order to define the
wrapping. Given that we do not use any wrapping, we can remove it from
the semantics.

In fact, it's probably better to decouple crash-freeness and predicate
satisfaction.

Here the new version of the semantics : $e \in \{x \mid p \} \iff e
\mbox{ diverges or } (p[e/x] \not \to^\star \{\bad,False\})$

But it leads to some tricky cases:

Assume a reasonable definition of $add$. Then, this contract does not
hold.
$$add ::: a:\{ x \mid True \} \to \{ y \mid True \} \to \{ z \mid z
\geq a \}$$

Why? Because now that we have removed crash-freeness, we can have $x$
and $y$ instantiated to $\bad$. And then, $ add~x~y \geq z $ becomes
$\bad$. Or the semantics say that the predicate should not evaluate to
$\bad$.

A possible fix is to provide, directly in the source language a $cf$
predicate, so that one can write contract like:

$$add ::: a:\{ x \mid cf(x) \} \to \{ y \mid cf(y) \} \to \{ z \mid z
\geq a \}$$

Note that in this setting $\{ x \mid True \} = \any$ and $\{ x \mid
cf(x) \} = \ok$.

\subsection{Dimitrios point}
Dimitrios isn't convinced that we keep the consistency of the whole
module when checking one function after the other.

Assume that we have a module that contains two functions defintion $f$
and $g$ and two contracts : $f ::: Any -> Ok$ and $g ::: Any ->
Ok$. We assume that those contract do not hold, for example if $f$ is
$not$ and $g$ is $id$.

Now, we want to check $f$'s contract. We take out $f$'s contract,
translate the function definition for $f$ and $g$ and add the contract
satisfation formulae for $g$'s contract.

But, given that $g$'s contract does not hold, we can derive $\bot$ and
then prove that $f$'s contract hold.

Now the user is happy, and does the same thing for $g$'s contract. It
holds too, given that we can also derive $\bot$ for $f$'s contract.

Finally, the user is really happy, but in fact he has proved nothing.

I think that it's not an actual issue, if we only give to equinox
useful formulae. Ie, if $f$ does not use $g$ we do not include $g$ in
the translation. Obviously it slows down the proof and, breaks the
consistency.

Consider the following graph, an arrow from $a$ to $b$ meaning that
$b$ uses $a$ in its definition.

\begin{center}
\includegraphics[scale=0.5]{flow.png}
\end{center}

We (D and I) believe that if we check a function by only giving the
formulae coming from the ancestors, then we do not break consistency.
Eg, to check $a$'s contract we need to give the contract and
definition for $f$ and $g$ too, but not for $h$.
